### 2021-07-06 rad-icfp-2021/correctness.tex
rewriting UNF section
writing section 5.4

We can interpret Source in $\CartSp$ as follows.
    \begin{tabular}{r c l}
    $\seml \reals \semr$ &\defeq& $\RR$ \\
    $\seml$!T1xT2!$\semr$ &\defeq& $\seml$!T1!$\semr \times\seml$!T2!$\semr$ \\
    $\seml \Array{\reals}{n}\semr$ &\defeq& $\Prod_{1\leq i \leq n} \RR$ 
    \end{tabular}
A context $\Gamma=\{x_1:A_1,\ldots,x_n:A_n\}$ is interpreted as the product $\Prod_{1\leq i \leq n}\sem{A_i}$.
Well typed terms $\Gamma\vdash$!e: !$A$ are interpreted as functions $\sem{\Gamma}\to\sem{A}$. This is
routinely defined by induction on the structure of typing derivations. Constants
!c! $\reals$ are interpreted as constant functions; and the primitives operations !op1,op2!
are interpreted by composing with the corresponding functions, which are smooth. 
Variables are interpreted as $\sem{x_i}(\rho)\defeq \rho_i$. 
The remaining constructions are interpreted as follows.
\begin{tabular}{r c l}
    $\seml$!map2 (x,y.e1) e2 e3!$\semr$ &=& TODO\\
    $\seml$!reduce (x,y.e1) e2 e3!$\semr$ &=& TODO
\end{tabular}
\end{definition}

@@

First, we start from the correctness of $D$, which is easy to establish, 
and then propagate this information to Source and Target.

\begin{proposition}[Correctness $D$]
    TODO
\end{proposition}

Recall that the intuition from Source UNF is that it consists of terms of Source that also return their context.
From there, the intuition for the Jacobian of a primitive in Target UNF is that it should be the Jacobian of
the corresponding term in Target.

TODO: I don't want to talk too much about the cotangent bundle functor, but to phrase correctness properties clearly I might have to.

\begin{proposition}[The transpose Jacobian of map2 and reduce are correct.]
    TODO
\end{proposition}

From this, we can now deduce that the composite transformation doing UNF then $D$ and then $UNF^-1$ is correct
in the sense that it produces a term that computes the gradient of the original term.

\begin{theorem}[UNF;D;$UNF^-1$ correct]
    TODO
\end{theorem}

By inspection of what that composition actually gives on terms of Source, 
we show that the previous composition of transformation indeed gives the transformation 
$\directD{\rho}{\Gamma}{Y}$ from Section~\ref{sub:Macro for pure reverse mode transformation}.

\begin{proposition}[UNF;D;$UNF^-1$=D]
    TODO
\end{proposition}

Without any more work, the previous claims have shown that 
$\directD{\rho}{\Gamma}{Y}$ is correct.

\begin{theorem}[D is correct]
    TODO
\end{theorem}

>>>>

We can interpret Source in $\CartSp$ as follows.
    \begin{tabular}{r c l}
    $\seml \reals \semr$ & $\defeq$& $\RR$ \\
    $\seml$!T1xT2!$\semr$ & $\defeq$& $\seml$!T1!$\semr \times\seml$!T2!$\semr$ \\
    $\seml \Array{\reals}{n}\semr$ & $\defeq$ & $\prod_{1\leq i \leq n} \RR$ 
    \end{tabular}
A context $\Gamma=\{x_1:A_1,\ldots,x_n:A_n\}$ is interpreted as the product $\prod_{1\leq i \leq n}\sem{A_i}$.
Well typed terms $\Gamma\vdash$!e: !$A$ are interpreted as functions $\sem{\Gamma}\to\sem{A}$. This is
routinely defined by induction on the structure of typing derivations. Constants
!c! $\reals$ are interpreted as constant functions; and the primitives operations !op1,op2!
are interpreted by composing with the corresponding functions, which are smooth. 
Variables are interpreted as $\sem{x_i}(\rho)\defeq \rho_i$. 
The remaining constructions are interpreted as follows.
\begin{tabular}{r c l}
    $\seml$!map2 (x,y.e1) e2 e3!$\semr$ &=& TODO\\
    $\seml$!reduce (x,y.e1) e2 e3!$\semr$ &=& TODO
\end{tabular}
\end{definition}

@@

First, we start from the correctness of $\Dsynrevsymbol_{\rho}$:Source UNF $\to$ Target UNF, which is easy to establish, 
and then propagate this information to Source and Target via the $\UNFSymbol, \UNFSymbol^{-1}$ transformations.

\begin{proposition}[Correctness $\Dsynrevsymbol_{\rho}$]
    For every term !$\reals^{\times n} \vdash$ e: $\reals^{\times n+1}$! in Source UNF,\\ 
    $\pi_2 \seml \Dsynrevsymbol_{\rho}$!e!$\semr(x_1,\ldots x_n,Id_{\RR^n})$=$J^T_{(x_1,\ldots x_n)}\seml$!e!$\semr$.
\end{proposition}

This is routinely proved by induction as the language is first-order. 
This uses the fact that for every primitive constant $A$, $\sem{J^TA}=J^T\sem{A}$.

Recall that the intuition from Source UNF is that it consists of terms of Source that also return their context.
From there, the intuition for the Jacobian of a primitive in Target UNF is that it should be the Jacobian of
the corresponding term in Target. This is easily checked for scalar operations.

\begin{proposition}
    \begin{center}
\begin{tabular}{r c l}
    !$\seml\UNFSymbol^{-1}(J^T$map2$_{x,y.e})\semr$! &=& !$J^T\seml\UNFSymbol^{-1}($map2$_{x,y.e})\semr$!\\
    !$\seml\UNFSymbol^{-1}(J^T$reduce$_{x,y.e_1;e_2})\semr$! &=& !$J^T\seml\UNFSymbol^{-1}($reduce$_{x,y.e_1;e_2})\semr$!
\end{tabular}
\end{center}
\end{proposition}

This is proved in Appendix \ref{sub:Reverse derivative of array operations}.

From this, we now deduce that the composite transformation $\UNFSymbol$, $\Dsynrevsymbol_{\rho}$, $\UNFSymbol^{-1}$ is correct
in the sense that it produces a term that computes the gradient of the original term.

\begin{proposition}
    If !x$_1:\reals$,$\ldots$,x$_n:\reals$ $\vdash$ e: $\reals$! then \\
    $\seml \pi_2 \UNFSymbol^{-1}(\Dsynrevsymbol_{\rho}(\UNFSymbol($!e!$)))\semr(x_1,\ldots x_n,Id_{\RR^n})$=$J^T_{(x_1,\ldots x_n)}$!<!$Id_{\RR^n}$,$\seml$!e!$\semr$!>!
\end{proposition}

\begin{proof}
    TODO
\end{proof}

By inspecting what that composition of transformations does on terms of Source, 
we show that this indeed gives computes the same as the transformation 
$\directD{\rho}{\Gamma}{Y}$ from Section~\ref{sub:Macro for pure reverse mode transformation}.

\begin{proposition}
    $\seml\UNFSymbol^{-1}(\Dsynrevsymbol_{\rho}(\UNFSymbol($!e!$)))\semr$ = $\seml\directD{\rho}{\Gamma}{Y}($!e!$)\semr$ 
\end{proposition}

\begin{proof}
    TODO
\end{proof}

Combining the the previous propositions, we have shown that $\directD{\rho}{\Gamma}{Y}$ is correct.

\begin{theorem}
    For every $\Gamma \vdash$ !e:!$\reals$, we have 
    $\sem{\grad_\Gamma e}= \grad_\Gamma\sem{e}$.
\end{theorem}
